An Introduction to Rocker: Docker Containers for R

### 2017-07-18 manuscript.Rmd
WIP

The `rstudio` image includes a lightweight, easy-to-use and
docker-friendly `init` system, s6 [@s6] for running persistant services,
including the RStudio server.  This system provides a convenient way for
downstream Dockerfile developers to add additional persistent services
(such as an ssh server) to a single container, or additional start-up
or shutdown scripts that should be run when an instance starts up or
shuts down.  The `rstudio` image uses such a start-up script to configure
user settings such as login password and permissions through environmental
variables at run time, see *Using Rocker Images* below.

@@

<!--
This stack also makes use of Docker metadata labels: <http://schema-label.org> metadata for `license`, `vcs-url` (GitHub repository), and `vendor`.  These can be altered or extended in downstream images. 
-->

@@

@@

- Portable: from laptop to cloud
- Sandboxed: experiment without touching local system
- Optimized: community-tweaked configurations
- Interfaces: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
- Extensible: interactively and persistent community extensions
- Versioned:

research use, instructional use

@@

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R envionment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers, or private or institutional server hardware.  Being
able to test code in a predictable, pre-configured R environment on a
local machine and then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such contanerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

The following code installs Docker on almost any Linux-based server and
launches a Rocker container providing the RStudio-server environment
over a web interface.

@@

After running these two lines, the user may log into the server merely
by pasting its IP address or DNS name into a browser and entering the
appropriate password.  This provides the user with a familiar,
interactive environment running on a remote machine while requiring
a minimum of expertise.

The user can select a Rocker
container which contains the required software pre-installed, such as
`rocker/tidyverse`, which provides the popular `tidyverse` suite of R
packages \citep{tidyverse}, or even use their own containers which can
extend the Rocker containers, as discussed below.  This workflow saves
considerable time relative to installing each package from scratch,
which can take over an hour to reconstruct the environment provided by
some rocker containers.  As many commericial cloud providers charge by
the hour, a fast and simple mechanism to deploy the desired environment
in the cloud is crucial to making this approach practical for more
exploratory analyses.

Unlike virtual machine images, multiple such Docker containers can
be run on the same server in the same way (though unique TCP ports
must be assigned to each if they are accessed through the RStudio web
interface). More importantly, it is much easier to test software locally
using the identical Docker container environment before deploying on
a remote server than it is to run a virtual machine such as an Amazon
Machine Image (AMI) locally, or to reproduce the runtime environment
identically on local and server systems.

@@

Rocker images may be used for quickly deploying R code on remote servers
without the use of the RStudio interface, running R in terminal mode
over an `ssh` connection. As with any interactive docker instance, users
should specify the interactive (`-i`) and terminal (`-t`, here combined
with interactive as `-ti`) flags, and specify the desired executable
environment (e.g. `R`, though other common options may be `Rscript` or
`bash`):

@@

While more detailed documentation on the use of Rocker images can be
found on the Rocker wiki, we provide a conceptual overview here which
illustrates common patterns and capabilities.

**Interactive shell**:

**RStudio instance**:

**Linking local volumes**: Docker

@@

**Root vs user instances**: By default, all rocker containers run as
root. This is consistent with standard Docker practice, allowing both
interactive users and downstream Dockerfiles to install software directly
without having to switch to root.  The main snag of this approach comes if
a user links a local volume and modifies files on that volume, which will
result in the local file being owned by `root` and not the default user.
To avoid this, specify the non-root user at runtime (username `docker`
on the `r-base` stack, or `rstudio` on images deriving from rstudio)
when running an interactive shell.  For users accessing R through RStudio,
this is not necessary.  The docker container will run as root, but a user
logs in through the RStudio server web interface as the non-root user
"rstudio", and thus any changes made to linked volumes will not alter file
permissions on the home directory.  For more details on user permissions,
see the Rocker wiki.

### Convenience

Installation of certain R packages can be challenging and non-intuitive,
particularly when those packages require specific external
dependencies. By running R through a Rocker image, users can deploy an
R environment where their desired packages and libraries have already
been installed and configured. As we discuss below, this configuration
benefits from the input of a large community of users, helping to
ensure better and more reliable performance. For instance, one example
of such a configuration is in building and testing R packages using
compiler sanatizers \citep{edd_sanitizers}, which requires R itself
be re-compiled with specialized settings not avaialable or familiar
to most R users.  Using the Rocker images `rocker/r-devel-san` and
`rocker/r-devel-ubsan-clang` provides a fast and simple way for users
to test

Many users are reluctant to upgrade their suite of installed packages,
which may break their existing code or even their R environment if the
installation goes poorly.  However, upgrading packages and/or the R
environment is often necessary to run analysis of a colleague or access
more recent methods. Using Rocker, users can run the latest or specific
earlier versions of R and R packages in an isolated environment without
impacting their local system.

### Extensible and Community-optimized images

Rocker images benefit from widespread use, testing and contributions from many members of the R community.

## Extending Rocker images

@@

### Reproducibility

\citep{Boettiger2015}

### Instruction

- Table of main repos, links to Docker-Hub & GitHub addresses

- list/table of topics on the wiki

>>>>

The `rstudio` image includes a lightweight, easy-to-use and
docker-friendly `init` system, s6 [@s6] for running persistant services,
including the RStudio server.  This system provides a convenient way for
downstream Dockerfile developers to add additional persistent services
(such as an ssh server) to a single container, or additional start-up
or shutdown scripts that should be run when an instance starts up or
shuts down.  The `rstudio` image uses such a start-up script to configure
user settings such as login password and permissions through environmental
variables at run time.  <!-- link wiki howto page? -->

@@

This stack also makes use of Docker metadata labels defined by
<http://schema-label.org>,
indicating image `license` (GPL-2.0), `vcs-url` (GitHub repository), and `vendor` (Rocker Project).  These metadata can be altered or extended in downstream images.

@@

<!-- 
Use cases: instruction, development, 
-->

@@

- **Portable**: from laptop to cloud
- **Interfaces**: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
- **Sandboxed**: experiment without touching local system
- **Optimized**: community-tweaked configurations
- **Extensible**: interactively and persistent community extensions
- **Versioned**: reproducible environments

@@

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R envionment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers [@jetstream], or private or institutional server hardware.  Being
able to test code in a predictable, pre-configured R environment on a
local machine and then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such contanerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

<!-- comment on use in instructional context -->
This portablility is similarly valuable in an instructional context. Requiring students
to install all necessary software on personal laptops can be particularly challenging
for short workshops, where download and installation time and troubleshooting across
heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
By deploying a Rocker image or Rocker-derived image (see extensibility) on a cloud machine,
an instructor can easily provide all students access to the pre-configured software environment
using only the browser on their laptops.  Unlike a vendor-specific virtual machine image (e.g. Amazon Machine Images, AMIs),
this approach does not preclude students being able to run the same software locally later on.
After installing Docker for their machine they too can run the course-specific software environment
easily on their own laptops and deploy it on other cloud providers.  This strategy has proven
effective in our own experience in both workshops and semester-length courses. Similar Docker-based
cloud deployments have been scaled to courses of 100s of students, e.g. at Duke [@Mine] and
UC Berkeley [@data8].

<!-- NOTE: Not sure if we should include such example commands, may be best left to the documentation? -->

For instance, the following code installs Docker on almost any Linux-based server and
launches a Rocker container providing the RStudio-server environment
over a web interface.

@@

Many academic and commerical cloud providers make it
possible to execute such code snippets when an instance is launched, without
ever needing to `ssh` into the machine.
After running these two lines, the user may log into the server merely
by pasting its IP address or DNS name into a browser and entering the
appropriate password.  This provides the user with a familiar,
interactive environment running on a remote machine while requiring
a minimum of expertise.

@@

An important aspect of the Rocker project design is the ability
for users to interact with the software on the container through either
an interactive shell session (such as the R shell or a bash shell), or
through a web browser accessing the RStudio\textsuperscript{\textregistered} Server
integrated development environment (IDE).  Traditional remote and high-performance
computing workflows for R users have usually required the use of `ssh` and 
a terminal only interface, posing a challenge for interactive graphics 
and a barrier to users unfamiliar with these tools and environments. Accessing
an RStudio instance through the browser removes these barriers.  Rocker
images include the RStudio-server software pre-installed and configured with
the explicit permission of RStudio Inc.

Users of the `rstudio` instances can access a root `bash` shell in an 
instance running the container using

```
docker exec -ti <container-id> bash
```

which can be useful for administrative tasks such as installing system dependencies.
All Rocker images can also be run as an interactive R, RScript or bash shell without running
RStudio, which can be useful for batch jobs or for anyone who prefers that environment.

<!--  Reserve these details for documentation, reference here? --> 
As with any interactive docker instance, users
should specify the interactive (`-i`) and terminal (`-t`, here combined
with interactive as `-ti`) flags, and specify the desired executable
environment (e.g. `R`, though other common options may be `Rscript` or
`bash`):

@@

**Linking local volumes**

@@

<!-- Blah blah blah ... link to documentation instead(?) -->
**Root vs user instances**: By default, all rocker containers run as
root. This is consistent with standard Docker practice, allowing both
interactive users and downstream Dockerfiles to install software directly
without having to switch to root.  The main snag of this approach comes if
a user links a local volume and modifies files on that volume, which will
result in the local file being owned by `root` and not the default user.
To avoid this, specify the non-root user at runtime (username `docker`
on the `r-base` stack, or `rstudio` on images deriving from rstudio)
when running an interactive shell.  For users accessing R through RStudio,
this is not necessary.  The docker container will run as root, but a user
logs in through the RStudio server web interface as the non-root user
"rstudio", and thus any changes made to linked volumes will not alter file
permissions on the home directory.  For more details on user permissions,
see the Rocker wiki.

### Sandboxed

Another feature of Rocker instances is the ability to provide a sandboxed 
environment, isolated from software and potentially from other data on
the machine.  Many users are reluctant to upgrade their suite of installed packages,
which may break their existing code or even their R environment if the
installation goes poorly.  However, upgrading packages and/or the R
environment is often necessary to run analysis of a colleague or access
more recent methods.  Rocker instances offer an easy solution. For 
instance, a user can run R code requiring the most recent versions of R
and related packages inside a Rocker container without having to upgrade their
local installations first.  Conversely, one could use Rocker to run code
on an older R release with prior versions of R packages, again without
having to make any alteration to one's local R install.  Another common
use case is to access an instance of R with support for particular
options such as using gcc or clang compiler sanatizers \citep{edd_sanitizers}.
These requires R itself be built with specialized settings that may not be
not available or familiar to many R users on their native system, but can
be easily deployed by pulling the Rocker images `rocker/r-devel-san` or
`rocker/r-devel-ubsan-clang`.

This sandboxing feature of Docker containers is also valuable in the 
remote computing context, where it can allow system administrators to
provide users with administrative (root) privilages to alter and install
software within a container without altering the computational environment
of other users on the remote machine.  Unlike traditional virtual machines,
these containers do not impose a large footprint of reserved resources --
a typical host can easily support 100s of containers [@docker-vs-vm].

<!-- volume linking isolation -->

### Community Optimized

Having a shared, transparent computational environment created by a publicly hosted, reproducible
recipe facilitates community input into configuration details. R and many of its packages
and related software can be configured with a wide range of options, compilers, different
linear-algebra libraries and so forth.  While this flexibility reflects varying needs,
many users rely on default settings which may be optimized more for simplicity of installation
than performance.  The Rocker recipes reflect significant community input on these choices,
as well as the considerable experience and expertise of the R Debian mantainer of 20 years
in what configuration options to use.  This helps create a more finely tuned, optimized reference
implementation of the R environment as well as a platform for comparing and discussing these 
issues which are often overlooked elsewhere.  Issues and Pull Requests on the Rocker 
repositories on GitHub attest to these discussions and improvements.
Widespread use of the Rocker image helps promote both testing of these choices and 
contributions further tweaking the configuration from many members of the R community.

### Versioned

[@Boettiger2015]

### Extensible

Rocker images are designed to provide

@@

## Conclusions
